Dependent Object Types

### 2012-08-30 doc/candidacy/report.tex
some corrections

\IEEEPARstart{T}{he} aim of this research is to propose a new sound
and tractable type-theoretic foundation of Scala~\cite{ScalaOverview} and languages like
it: the Dependent Object Types calculus (DOT). The properties we are
interested in modeling are Scala's path-dependent types and abstract
type members, as well as its mixture of nominal and structural typing
through the use of refinement types. Compared to previous
approaches~\cite{nuObj,FS}, we make no attempt to model inheritance or
mixing composition. Furthermore, the calculus is more normative than
descriptive, in that it does not precisely describe what's currently
in Scala.

@@

The terms in ${\nu}Obj$ denotes objects or classes. They consist of
variables $x$ denoting objects, selections $t.l$, object creations
\mbox{$\newin{x \larrowc t}{u}$} defining an object $x$ of class $t$
whose scope is the term $u$, class templates $\cls{x: S \sep \tup d}$,
and mixing compositions $t \withsub{S} u$ denoting a class of self
type $S$ combined from two classes $t$ and $u$. A value is a variable
or a class template. A path is a variable followed by a possibly empty
sequence of selections.

@@

Contrasting concrete type bindings $\tup d$ used in class template
terms with type declarations $\tup D$ used in record and class types,
the possible type binders in type declarations are type aliases $=$,
new types $\expand$ and abstract types $<:$. Only the first two are
also concrete type binders available in type bindings of class
templates. A consequence is that classes with abstract type members
cannot be instantiated, because each type declaration of the self type
needs to have a corresponding type binding when an object is
instantiated.

Mixing composition is not commutative. When expanding a type to a
record type, a compound type $R_1 \with R_2$ expands to a record type
containing the concatenation of the declarations in $R_1$ and $R_2$,
but if some label is defined in both $R_1$ and $R_2$, the definition
in $R_2$ overrides (and so must be more specific than) the definition
in $R_1$. Though this break of commutativity is justified from an
implementation standpoint, it is awkward from a typing standpoint: in
particular, least upper bounds and greatest lower bounds
do not always exist in Scala.

Let's now illustrate with a few examples the correspondences outlined
earlier. %TODO

>>>>

\IEEEPARstart{T}{he} aim of this research is to propose a new sound
and tractable type-theoretic foundation of Scala~\cite{ScalaOverview} and languages like
it: the Dependent Object Types calculus (DOT). The properties we are
interested in modeling are Scala's path-dependent types and abstract
type members, as well as its mixture of nominal and structural typing
through the use of refinement types. Compared to previous
approaches~\cite{nuObj,FS}, we make no attempt to model inheritance or
mixin composition. Furthermore, the calculus is more normative than
descriptive, in that it does not precisely describe what's currently
in Scala.

@@

The terms in ${\nu}Obj$ denotes objects or classes. They consist of
variables $x$ denoting objects, selections $t.l$, object creations
\mbox{$\newin{x \larrowc t}{u}$} defining an object $x$ of class $t$
whose scope is the term $u$, class templates $\cls{x: S \sep \tup d}$,
and mixin compositions $t \withsub{S} u$ denoting a class of self
type $S$ combined from two classes $t$ and $u$. A value is a variable
or a class template. A path is a variable followed by a possibly empty
sequence of selections.

@@

Contrasting concrete type bindings $\tup d$ used in class template
terms with type declarations $\tup D$ used in record and class types,
the possible type binders in type declarations are type aliases $=$,
new types $\expand$ and abstract types $<:$. Only the first two are
also concrete type binders available in type bindings of class
templates.

A class type $\cls{x: S \sep \tup D}$ contains as values classes that
instantiate to objects of (sub)type $\obj{x \sep \tup D}$. The
explicit self type $S$ may be different from $\obj{x \sep \tup
  D}$. Declarations in $S$ which are not in $\tup D$ play the role of
abstract members -- they must be defined by mixin composition in order
for the class to be instantiated.

Mixin composition is not commutative. When expanding a type to a
record type, a compound type $R_1 \with R_2$ expands to a record type
containing the concatenation of the declarations in $R_1$ and $R_2$,
but if some label is defined in both $R_1$ and $R_2$, the definition
in $R_2$ overrides (and so must be more specific than) the definition
in $R_1$. Though this break of commutativity is justified from an
implementation standpoint, it is awkward from a typing standpoint: in
particular, least upper bounds and greatest lower bounds
do not always exist in Scala.

Let's now illustrate with a few examples the correspondences outlined
earlier.

First, we show how to encode monomorphic functions into the
calculus. The paper goes further and generalizes the encoding to
system $F_{<:}$
